---
title: 'DATA 607: Week 3 Assignment'
author: "Michele Bradley"
date: "9/14/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Introductory Example

```{r}
library(stringr)
raw.data <-"555-1239Moe Szyslak(636) 555-0113Burns, C. Montgomery555-6542Rev. Timothy Lovejoy555 8904Ned Flanders636-555-3226Simpson, Homer5553642Dr. Julius Hibbert" 
names <-unlist(str_extract_all(raw.data, "[[:alpha:]., ]{2,}"))
names
```

### Question 3 

#### 3.a Rearrange the Vector so all elements conform to the standard first_name last_name

```{r}
first_name_extract <- unlist(str_extract_all(names, "\\w+\\s|[, ]\\s\\w+"))
first_name_split <- unlist(str_split(first_name_extract, ",[[:blank:]]{1}"))
first_names <- unlist(str_extract_all(first_name_split, "\\w+"))
first_names
```

```{r}
last_name_extract <- unlist(str_extract_all(names, "[^[:punct:]]\\s\\w+$|\\w+[,]"))
last_name_split <- unlist(str_split(last_name_extract, "[[:blank:]]{1}"))
last_names <- unlist(str_extract_all(last_name_split, "[[:alpha:]][[:alpha:]]+"))
last_names
```

```{r}
Simpsons_Characters <- data.frame(first_names, last_names)
Simpsons_Characters
```

#### 3.b Construct a logical vector indicating whether a character has a title 

```{r}
unlist(str_detect(names, "[A-Za-z][A-Za-z][.]"))

titled_name <- unlist(str_extract_all(names, "\\w+[.]\\s\\w+\\s\\w+"))
titled_name
```

#### 3.c Construct a logical vector indicating whether a character has a second name.

```{r}
unlist(str_detect(names, "\\s\\w[.]\\s\\w+"))
```

### Question 4
### Describe the types of strings that conform to the following regular expressions and construct an example that is matched by the regular expression

#### (a) [0-9]+\\$
#### Answer: A digit followed by a dollar sign

```{r}
sample4a <- c("456$", "87", "7802$", "hello", "telephone$", "45%", "0190$", "$67", "67$2")
str_detect(sample4a, "[0-9]+\\$")
```

#### (b) \\b[a-z]{1,4}\\b
#### Answer: A lowercase word between 1 and 4 characters long

```{r}
sample4b <- c("far", "May","vba3", "b", "hello")
str_detect(sample4b, "\\b[a-z]{1,4}\\b")
```

#### (c) .*?\\.txt$
#### Answer: Any combination of characters (or no characters) and matched at most once that is a .txt file

```{r}
sample4c <- c("hello", "dsfjh", "hello.txt", "dsfjh.txt", "123.txt", "%^*(&%^.txt", ".txt")
str_detect(sample4c, ".*?\\.txt$")
```

#### (d) \\d{2}/\\d{2}/\\d{4}
#### Answer: Two digits / Two Digits / Four Digits

```{r}
sample4d <- c("67/89/0984", "445/2346/982357", "53/75/1893", "04.24.1995", "04/24/1995")
str_detect(sample4d, "\\d{2}/\\d{2}/\\d{4}")
```

#### (e) <(.+?)>.+?</\\1>
#### Answer: Any mix of characters enclosed by <></>

```{r}
sample4e <- c("<header>Hello World</header>", "<header>Hello World<header>")
str_detect(sample4e, "<(.+?)>.+?</\\1>")
```

### Question 9 
#### The following code hides a secret message. Crack it with R and regular expressions.

```{r}
secret_message <- "clcopCow1zmstc0d87wnkig7OvdicpNuggvhryn92Gjuwczi8hqrfpRxs5Aj5dwpn0TanwoUwisdij7Lj8kpf03AT5Idr3coc0bt7yczjatOaootj55t3Nj3ne6c4Sfek.r1w1YwwojigOd6vrfUrbz2.2bkAnbhzgv4R9i05zEcrop.wAgnb.SqoU65fPa1otfb7wEm24k6t3sR9zqe5fy89n6Nd5t9kc4fE905gmc4Rgxo5nhDk!gr"

break_message = unlist(str_extract_all(secret_message, "\\w+[[:punct:]]"))
message = unlist(str_extract_all(break_message, "[[:upper:]]|[[:punct:]]"))
message_readable = paste(message, collapse="")
message_readable
```




